---
title: "Reproducible Research: Peer Assessment 1"
By: Trupti B Palande
html_document:
keep_md: true
---

## Loading and preprocessing the data
1.Load data from activity.csv file assumming it is extracted in data folder under working directory Ex : D:\RepData_PeerAssessment1\Data


```{r}
##set working directory
##setwd("D:/Assignments/DataScienceCoursera/Reproducible research/RepData_PeerAssessment1")
activity <- read.csv("./Data/activity.csv", colClasses = c("integer", "Date", "factor"))
class(activity)
head(activity)

```
2. process the data by removing records with NA data

```{r}
activitynoNA <- na.omit(activity)
rownames(activitynoNA) <- 1:nrow(activitynoNA)
head(activitynoNA)
dim(activitynoNA)
```


## What is mean total number of steps taken per day?

a histogram of the total number of steps taken each day

```{r}
library(ggplot2)
library(plyr)
library(reshape2)

## A histogram of the total number of steps taken each day
ggplot(activitynoNA, aes(date, steps)) + geom_bar(stat = "identity", colour = "steelblue", fill = "steelblue", width = 0.7) + labs(title = "Histogram of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")

## Calculate the mean total number of steps taken per day
totalSteps <- aggregate(activitynoNA$steps, list(Date = activitynoNA$date), FUN = "sum")$x
oldMean<-mean(totalSteps)

## Calculate the Median total number of steps taken per day
oldMedian<-median(totalSteps)

oldMean
oldMedian
```
** Mean total number of steps per day: 10766.19 **
**	Median total number of steps per day: 10765 **

## What is the average daily activity pattern?

* Time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
## calculte average steps taken across all days over 5 minutes interval
avgSteps <- aggregate(activitynoNA$steps, list(interval = as.numeric(as.character(activitynoNA$interval))), FUN = "mean")
names(avgSteps)[2] <- "meanOfSteps"

##plot time series plot 
ggplot(avgSteps, aes(interval, meanOfSteps)) + geom_line(color = "steelblue", size = 0.8) + labs(title = "Time Series Plot of the 5-minute Interval", x = "5-minute intervals", y = "Average Number of Steps Taken")
```


*Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}

index <- which.max(avgSteps[,2])
interval <- avgSteps[index,1]
maxSteps <- avgSteps[index,2]

interval
maxSteps
```


**  interval is : 835  **
**  max Steps are: 206.1698 **

## Imputing missing values

* total records with missing values
```{r}
miss<-sum(is.na(activity$steps))
miss
```
** total number of missing values in the dataset: 2304 **

** For imputing missing values we have two options either impute missing values with the **
** mean of corresponding 5-minute interval or the mean/median for that day. Will impute **
** missing values  with the mean for that 5-minute interval. **

```{r}
newActivity <- activity
for (i in 1:nrow(newActivity)) {
  if (is.na(newActivity$steps[i])) {
    newActivity$steps[i] <- avgSteps[which(newActivity$interval[i] == avgSteps$interval), ]$meanOfSteps
  }
}

head(newActivity)
sum(is.na(newActivity))

```

** Plot histogram of the total number of steps taken each day and Calculate new mean and  **  
** median after imputing missing values to see difference **
```{r}
## plot histogram
ggplot(newActivity, aes(date, steps)) + geom_bar(stat = "identity", colour = "steelblue", fill = "steelblue", width = 0.7) + labs(title = "Histogram of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")

## calculate new mean and median
newtotalSteps <- aggregate(newActivity$steps, list(Date = newActivity$date), FUN = "sum")$x
newMean<-mean(newtotalSteps)
newMedian<-median(newtotalSteps)
newMean
newMedian
```
*  Mean total number of steps per day after imputing missing values : 10766.19 
*  Median total number of steps per day after imputing missing values: 10766.19



```{r}
## calculate difference between old and new values of mean and median
diffMean<-newMean - oldMean
diffMedian<-newMedian - oldMedian

diffMean
diffMedian
```
Here we see that the impact of replacing the missing values is negligible on mean and median of the total steps taken per day. Mean is almost same and Median is off by 1 unit from the previous Median.


## Are there differences in activity patterns between weekdays and weekends?

* Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}

library(lattice)

actdiff <- newActivity

##add another column day specifying weekday as factor
actdiff$day <- factor(format(as.Date(newActivity$date), "%A"))

##Create the "weekday" and "weekend" levels.
levels(actdiff$day) <- list(weekday = c("Monday","Tuesday","Wednesday","Thursday","Friday"), weekend=c("Saturday","Sunday"))

## calculte average steps taken across all days based on "interval" and "day"
actdiffsteps <- aggregate(actdiff$steps, by=list(actdiff$interval, actdiff$day), mean)

##Assign the names of the columns
names(actdiffsteps) <- c("Interval", "Day", "Steps") 

## plot Lattice graph of weekdays and weekends for comparing activity pattern
xyplot(Steps ~ Interval | Day, data=actdiffsteps, layout=c(1,2), type="l" )

head(actdiffsteps)
levels(actdiffsteps$Day)

```
** There are minor differences in the pattern between weekends and weekdays. **
** Peak activity is greater on weekdays compared to weekends, however, **
** step activity is more equally distributed throughout the day on weekends **
** compared to weekdays, **
